Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea

### 2023-02-08 analysis/paper/paper.qmd
Done with results!

abstract: |
  The onset of the Late Paleolithic period in Korea, represented by the appearance of stemmed points and blades, was a key event in the dispersal of modern humans in Northeast Asia. Previous studies have mainly focussed on possible origin locations of these new technologies. The specific cultural processes of the appearance stemmed points and blades has rarely been considered. We investigate the cultural processes by applying a cultural transmission framework to investigate the social contexts of the emergence of these new technologies. Our main question is: what was the dominant mode of cultural transmission during this time of technological innovation in the Korean Late Paleolithic? Following Bettinger and Eerkens [-@bettinger1999point], we evaluated models of guided variation and indirect bias using data from Korean assemblages containing stemmed points. To evaluate these models and understand the transmission processes, we computed correlation coefficients and coefficients of variation (CV). We found that information about the new technology was likely transmitted via selective combinations of guided variation and indirect bias over time. Some attributes including length and width were transmitted with less variation while other attributes appear to have more variation. Our results suggest that the dominant mode of cultural transmission for the earliest stemmed points was guided variation and then indirect bias was adapted to carry out the technological change. We assume that individuals or groups developed stemmed points by experimenting with existing blade technologies and then copied crucial part of a successful model to ensure the quality to optimize tool usage. As a result, the shape of stemmed points became more standardized among their social groups.

@@

## Correlation coefficients and Variation Over Time

@@

The correlations between attributes became stronger over time as we can clearly see in the the Panel D of @fig-corr-analysis. The median for Phase 2 is much lower than `r threshold_value_cor`, which @bettinger1999point interpreted as a threshold to evaluate transmission biases, while the Phase 3 median is close to the threshold. Based on this results, we assume that the earlier transmission for stemmed points was more likely the results of guided variation. On the other hand, Phase 3 could be more associated with indirect bias.

@@

```
## Correlation between assemblages

@@

## Coefficient of variation and Variation over time

@@

Our results show that CV values for all attributes are distributed from `r min_cv` to `r max_cv` (@fig-plot-cv-all-attributes). Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV for the Korean artifacts is higher. Unlike the distribution of the linear measurements, the CVs of body length and maximum length have the lowest values. The lower CV of maximum length indicates they are highly standardized with narrowest confidence intervals. Tang and stem related attributes (e.g. SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes Among all the attributes, only maximum length has the CV values lower than 25, which suggests the transmission of this attribute was more influenced by indirect bias than the others.

@@

@@

Following the previous studies of [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we only compared CV values of attributes, body length, maximum length, mid width, and tang width. @fig-plot-cv-four-assemblage shows that CV values for the four assemblages are mostly slightly below `r threshold_value`. Among the assemblages, SYG1_2 (n=7) has higher CV values as well as the widest ranges of confidence intervals, perhaps due to its small sample size, relative to the other assemblages compared here. Overall we see only subtle changes in CV values from Phase 2 to Phase 3 in @fig-plot-cv-four-assemblage. This suggests that the complex directional trends in @fig-plot-cv-over-time might be best interpreted as noise in an overall signal of guided variation, rather than substantial changes in the type of bias dominating cultural transmission.

@@

## Variation Between sites

We examined the CV values per site to address our research question about the cultural transmission over space. We excluded assemblages that contain less than five points. @fig-plot-cv-by-site shows higher CV values for stem length and stem width, indicating high variability at the stem of the points from these sites. Overall size variability is similar between these three assemblages, suggesting a consistent approach to constraining shape variation over space, rather than site-specific variances.

```{r}
#| label: fig-plot-cv-by-site
#| fig-cap: CV values per site for each site has more than five stemmed points. The points represent CV values for each attribute. The vertical lines are confidence intervals.
#| fig-width:   8.5
#| fig-height:  4

cv_by_site_df  <-
  df_full_sitename %>%
  dplyr::select(-SPstage.Stage, 
                -lat_dd, 
                -long_dd, 
                -sitename, 
                -SPstage.Raw_material) %>%
  group_by(full_sitename) %>%
  add_tally() %>%
  dplyr::filter( n > 5) %>%
  dplyr::select(-n) %>%
  pivot_longer(-full_sitename,
               names_to = "variable",
               values_to = "value") %>%
  group_by(full_sitename, variable) %>%
  dplyr::summarise(cv_by_site =   sharma_cv(value),
            cv_low_sharma =  sharma_int_low(value),
            cv_high_sharma = sharma_int_high(value))

## CV for per each site (n=X) with the full site name : label

cv_by_full_site_df_label <-
  df_full_sitename %>%
  dplyr::select(-SPstage.Stage) %>%
  group_by(full_sitename) %>%
  add_tally() %>%
  dplyr::filter( n > 1) %>%
  mutate(label = as.factor(paste0(full_sitename,' (N = ', n, ")"))) %>%
  dplyr::select(-n) %>%
  nest(-full_sitename, -label) %>%
  dplyr::mutate(cv_by_site = purrr::map(data, ~map_df(.x, mccv))) %>%
  dplyr::select(-cv_by_site)

## add label to main dataframe

cv_plot_site_label <- cv_by_site_df %>%
  left_join(cv_by_full_site_df_label) %>%
  dplyr::select(- data,-full_sitename)

## facet plot

cv_plot_site_label %>%
  ggplot(aes(x = variable,
             y = cv_by_site)) +
  geom_point() +
  geom_errorbar(aes(ymin = cv_low_sharma,
                    ymax = cv_high_sharma),
                width = 0.2) +
  xlab("Attribute") +
  ylab("Coefficient of Variation on Attributes") +
    # draw a line to show our threshold value
  geom_hline(yintercept = 25, colour = "red") +
  facet_wrap( ~ label, scales = "free_y") +
  theme(legend.position = "none") +
  coord_cartesian(ylim = c(0,70)) +
  theme_bw() +
  theme(strip.text.x = element_text(size = 6.5))
```

@@

# Discussion {.unnumbered}

>>>>

abstract: |
  The onset of the Late Paleolithic period in Korea, represented by the appearance of stemmed points and blades, was a key event in the dispersal of modern humans in Northeast Asia. Previous studies have mainly focussed on possible origin locations of these new technologies. The specific cultural processes of the appearance stemmed points and blades has rarely been considered. We investigate the cultural processes by applying a cultural transmission framework to investigate the social contexts of the emergence of these new technologies. Our main question is: what was the dominant mode of cultural transmission during this time of technological innovation in the Korean Late Paleolithic? Following Bettinger and Eerkens [-@bettinger1999point], we evaluated models of guided variation and indirect bias using data from Korean assemblages containing stemmed points. To evaluate these models and understand the transmission processes, we computed correlation coefficients and coefficients of variation (CV). We found that information about the new technology was likely transmitted via guided variation with small impact of indirect bias. Some attributes including length and width were transmitted with less variation while other attributes appear to have more variation. Our results suggest that the dominant mode of cultural transmission for the earliest stemmed points was guided variation and then indirect bias was adapted to carry out the technological change. We assume that individuals or groups developed stemmed points by experimenting with existing blade technologies and then copied crucial part of a successful model to ensure the quality to optimize tool usage. As a result, the shape of stemmed points became more standardized among their social groups.

@@

## Correlation coefficients

@@

The correlations between attributes became stronger over time as we can clearly see in the Panel D of @fig-corr-analysis. The median for Phase 2 is much lower than `r threshold_value_cor`, which @bettinger1999point interpreted as a threshold to evaluate transmission biases, while the Phase 3 median is close to the threshold. Based on this results, we assume that the earlier transmission for stemmed points was more likely the results of guided variation. On the other hand, Phase 3 could be more associated with indirect bias.

@@

@@

## Coefficient of variation

@@

Panel B of @fig-plot-cv-all-attributes shows that CV values for all attributes are distributed from `r min_cv` to `r max_cv`. Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV for the Korean artifacts is higher. Unlike the distribution of the linear measurements, the CVs of body length and maximum length have the lowest values. The lower CV of maximum length indicates they are highly standardized with narrowest confidence intervals. Tang and stem related attributes (e.g. SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes. Overall, our results suggests that transmission of stemmed points could be more influenced by guided variation.

@@

Similar to Panel D of @fig-corr-analysis, we divided the stemmed points into three chronological phases and computed CV to examine temporal patterns in the modes of cultural transmission. Again, we excluded stemmed points for Phase 1 due to the limited number of artifacts.

@@

Following the previous studies of [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we only compared CV values of attributes, body length, maximum length, mid width, and tang width for the assemblages that have more than five stemmed points. @fig-plot-cv-four-assemblage shows that CV values for the four assemblages are mostly slightly below `r threshold_value`. Among the assemblages, SYG1_2 has higher CV values as well as the widest ranges of confidence intervals. It is interesting because SYG1_2 has the strong correlation between attributes (@fig-plot-cor-four-assemblage). Perhaps it could be related to its small sample size, relative to the other assemblages compared here. Overall we see only subtle changes in CV values from Phase 2 to Phase 3 in @fig-plot-cv-four-assemblage. This suggests that the complex directional trends in @fig-plot-cv-over-time might be best interpreted as noise in an overall signal of guided variation, rather than substantial changes in the type of bias dominating cultural transmission.

@@

@@

# Discussion
